---
title: "Introduction to Programming in R"
author: "[Pedro Cardoso-Leite](pedro.cardosoleite@uni.lu)"
title-block-banner: true
title-block-style: plain
affiliation-title: "University of Luxembourg" 
format: 
  revealjs:
    theme: [default, ../../../assets/revealjs_themes/teaching.scss]
    css: ../../../assets/revealjs_themes/style.css
    incremental: true
    chalkboard: true
    logo: "https://xcit.org/images/xcit/logofinaline.png"
    width: 1600
    height: 900
    pointer:
      color: "#db0d23"
      pointerSize: 24
    revealjs-plugins:
      - pointer
editor: visual
---

# The basics of programming (in R) {#color-slide background-image="../../../assets/background_images/eniac-programming.png"}

```{css, echo=FALSE}
#color-slide, 
#color-slide h1,h2 {
 color: white;
}
```

## ABC vs Writing a book

-   Fundamental skills

-   Specialized skills

-   Workflow, methodologies, tools

## Key concepts

1.  values / data (i.e., the content, e.g., `"hello"`)
2.  variables (data containers, e.g., `x <- 5`)
3.  loops (e.g., "repeat operation 1000 times")
4.  control flow (e.g., "if x then y else z")
5.  functions (i.e., encapsulate transformations)

## This is similar to Math!

-   f(x): 2\*x + 1
-   f(5)
-   z = 3; f(z)
-   g(x): x if x \> 0, else 0

## Touchtyping for coders

Coding (i.e., writing code) requires using characters you might not be used to typing. Take a couple of minutes to figure out how to type the following characters on your keyboard:


::::: {.columns}
::: incremental

:::: {.column width='50%'}
1.  ( ) \[\] { } round, square and curly brackets
2.  " ' double and single quote
3.  \- \_ hyphen and underscore
4.  \< \> smaller and greater than
5.  \| pipe
:::: 

:::: {.column width='50%'}
6.  \` back tick
7.  \\ / backward and forward slash
8.  \$ dollar sign
9.  \~ tilde
10. \# pound
11. & and operator
:::: 
:::
:::::



## Touchtyping

It is worth learning to touch type! 

There are many free online resources:

 - https://www.typingclub.com/
 - https://www.keybr.com/
 - https://monkeytype.com/
 - https://www.speedcoder.net/
 - https://typing.io/

::: incremental
Try
:::



## Values / data types

### Atomic values

-   Logical (boolean)
-   Integer
-   Numeric (double)
-   Character / String
-   ...

Exercise: find examples of values for each of these types.

## Values / data types

### Data structures

Atomic values can be organized into more complex data structures:

| dimensions | homogenous    | heterogenous |
|------------|---------------|--------------|
| 1          | atomic vector | list         |
| 2          | matrix        | data frame   |
| n          | array         |              |

: Data structures

## Exercise: what data structures are these?

Example 1:

```{r}
head(mtcars)
```

Example 2:

`("a", 12, TRUE)`

Example 3:

```{r}
round(matrix(data = rnorm(9), nrow = 3),2)
```

Example 4:

```{r}
"hello world!"

```

## Variables / containers

We can store values (and functions and other stuff) in labeled containers called variables!

Example:

`x <- 5`

This is read as:

-   "x gets 5"
-   "we *assign* the value 5 to variable x"

Note: most programming languages use "=" instead.

## Variables / containers

We can store more complex data structures:

`my_vector <- c(1,3,5,7)`

as well as transformations (more on this later):

`my_function <- function(x){2*x+1}`

## How to name variables?

### Requirements

-   must start with a letter or a period (followed by letter)
-   reserved words (e.g., `TRUE`, `if`, `function`) are forbidden
-   can't contain spaces (e.g., `my variable` vs `my_variable`)

### Recommendations

-   only use lower case letters, digits and underscore
-   use meaningful names
-   clarity beats brevity

## How to name variables?

Exercise: Find a good name for the variable `___`?

``` r        
___ <- "John"

if ( ___ == "John"){
  print("Hi John! Nice to see you again.)
} else {
  print(paste("Hi", ___, ". It looks like you are new to this course. Welcome!"))
}
```

# Practice!

## Let's try swirl! {.center}

**course**: R Programming Basic Building Blocks

**chapter**: Basic Building Blocks




## Working on a project

- R console
- R scripts (*.R)
- R + quarto files (*.qmd)
- R projects (*.Rproj)





## Functions

## Functions: Example 1

Calculate the mean of the following values:

```{r}
#| echo: true

age <- c(23, 26, 37, 45, 63, NA, 52)

```

Note: - getting help - argument, parameters - different functions expect different types of input - function outputs

### Functions: Example 2

calculate the correlation between age and weight

```{r}
#| echo: true
age <- c(23, 26, 37, 45, 63, NA, 52)
weight <- c(70, 64, 80, 63, 92, 102, 77)


```

## Create your own functions

To create your own functions use the following template:

```{r}
#| echo: true
#| eval: false
#| 
my_function_name <- function(input_1, input_2) {
  
  # --- write your code here --
  # example:
  output <- input_1 + input_2
  
  # return the output
  output
  
}

```

## Functions

-   Naming functions
-   namespace (e.g., `mean()` vs `base::mean()`)

## Exercise: BMI calculator (part 1)

Write a function that computes the [body mass index](https://www.nhlbi.nih.gov/health/educational/lose_wt/BMI/bmicalc.htm) (in an R script!)

$$ bmi = \frac{weight}{height²} $$

$weight$ in kg and $height$ in m

## Exercise: BMI calculator (part 2)

Change your function so it prints the text:

> "Your height is \_\_\_ m and your weight is \_\_\_ kg:

> Your BMI is \_\_\_."

## Exercise: BMI calculator (part 3)

Change your function so it prints the text:

> "Your height is \_\_\_ m and your weight is \_\_\_ kg:

> Your BMI is \_\_\_."

> Your BMI-category is \_\_\_."

## "Sourcing" code